---
title: "Homework 4 - Hodge Podge"
author: "Jason M. Pattison, ST 588-651 Summer 1 2024"
format: html
editor: visual
---

# Homework 4 Assignment Objectives

Task 1 for this homework assignment is practice with answering the types of questions that will be asked on the next exam.

Task 2 of the assignment involves writing R functions manually create functions that generate the Root Mean Square Error (RMSE) and Mean Absolute Deviation (MAE) given a set of response values.

Task 3 of the assignment focuses on querying an API and writing a function that queries the API and

## Task 1: Conceptual Questions

1.  What is the purpose of the lapply() function? What is the equivalent purrr function?

    > The purpose `lapply()` is to apply a function to all variables in a data frame. `lapply`treats the variables in data frames as a list, and applies a called function to all objects in the list. The equivalent purrr function to`lapply()`is`map()`.

2.  Suppose we have a list called my_list. Each element of the list is a numeric data frame (all columns are numeric). We want use lapply() to run the code cor(numeric_matrix, method = "kendall") on each element of the list. Write code to do this below! (I’m really trying to ask you how you specify method = "kendall" when calling lapply())

    > lapply(my_list, cor(numeric_matrix, method = "kendall"))

3.  What are two advantages of using purrr functions instead of the BaseR apply family?

    > `map()` is more consistent and has additional helper funcitons for use in analysis

4.  What is a side-effect function?

    > A side-effect funciton is one where the state of the data is changed while returning a value. Some examples are `print()` and `plot()` where the function takes the data and creates the different outputs.

5.  Why can you name a variable sd in a function and not cause any issues with the sd function?

    > The sd() function is global throughout R (via R packages). When use inside of a function, `sd` is a temporary variable that is local to the funciton it is in. Once the functioncode is finished running, the variable sd does not exist.

## Task 2: Writing R Functions

Step 1 of this task is to write a basic function called `getRMSE()` that takes in a vector of responses, a vector of predictions, and outputs the RMSE. 

```{r write the RMSE function}

# name the function "getRMSE", define "x" as our response values and "y" as our predicted values. 

getRMSE <- function(x, y) {


# generate the loop required to subtract each predicted value from the corresponding response value.

  for (i in x) {  


# subtract the predicted values from the observed values, square the difference, sum the values, establish the mean of the sum, then take the square root of the mean. 
# The order of which variable is subtracted from which doesn't matter due to squaring the difference for each iteration of the loop. 
    
  diff_sq <- ((x - y)**2)

  } 

  RMSE <- sqrt(mean(sum(diff_sq)))

    
  # output the result
  RMSE
}

```

2. Run the code provided to create vectors of response values and predicted values. Use these vectors to test `getRMSE`.When writing the function, we want the `mean()` to replace any NA values in the response

```{r testing the RMSE function}

set.seed(10)
n <- 100
x <- runif(n)
resp <- 3 + 10*x + rnorm(n)
pred <- predict(lm(resp ~ x), data.frame(x))


getRMSE <- function(x, y) {
  for (i in x) {  
    diff_sq <- ((x - y)**2)
  } 

  RMSE <- sqrt(mean(sum(diff_sq)))

    
  # output the result
  RMSE
}

getRMSE(resp, pred)

```
Now that we have successfully tested our function, we are going to replace two response values with missing values using (NA_real_), then test `getRMSE` function with and without specifying behavior to deal with missing values. 

First we will test getRMSE without specifying behavior to deal with the missing values. We will save the results of this test as `RMSE_NA_without_specs`. 

```{r replace response values, run getRMSE}

# replace two response values, create new response vector so as to not overwrite original data. 

replace_values <- c(resp[1], resp[100])

resp_2 <- replace(resp, resp %in% replace_values, NA_real_)


#test getRMSE without specifying behavior

RMSE_NA_without_specs <- getRMSE(resp_2, pred)

```

Running the new vector through `getRMSE` produces the result of "NA". This is due to the two NA values in the response variable not being something we can compute. 

To correct this, we will add code to `getRMSE` that will replace any NA values in an analyzed vector with the mean of the vector's other values. We will call the updated function `getRMSE_2`. 

```{r}

#update getRMSE to account for missing data in response variable, call the updated function getRMSE_2

getRMSE_2 <- function(x, y) {
  x <- replace(x, is.na(x), mean(x, na.rm = TRUE))
  for (i in x) {  
    diff_sq <- ((x - y)**2)
  } 

  RMSE <- sqrt(mean(sum(diff_sq)))

    
  # output the result
  RMSE
}

getRMSE_2(resp_2, pred)


```


## Task 3: Querying an API and a Tidy-Style Function
